The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-10-04 analysis/paper/_main.Rmd
formatting to submit

---
title: "The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
author:
  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-humanities-author-date.csl"
abstract: |
  Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions. Despite this, even after one century of being identified there is still no consensus on how to define, analyze or even interpret these tools. In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale. In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi. Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts. Our second goal was to critically evaluate the role of these artifacts within western Iberia’s Upper Paleolithic.
  Our results showed that functional identification of scaled pieces is still not clear. By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high. Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns. While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.

@@

library(knitr)
library(tidyverse)
library(broom)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)
library(sjstats)
library(powerAnalysis)
library(epitools)
#library(ScaledPiecesVB)

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of portuguese paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces. Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g. @zilhao_o_1997] or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997] for example argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.

@@

Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g. @bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018]. This can be explained, partially, by raw material availability. Most raw materials were procured locally, or regionally from deposits located at no more than 20 km away from the site [@bicho_paleolithique_2010]. 
Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production. Other raw materials can be found at the site, but much more restricted, both diachronically and functionally, within each techno-complex. Schist, for instance, shows up in some occupations almost exclusively connected to mobile art. Dolerite was only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].

@@

We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends. These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed. As will become clear from our results some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces. This detail has been previously referred by Villa et al. [-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.

Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes. In the first group, a series of variables traditionally used in lithic studies [e.g. @andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (see Supplementary Materials for full list of variables), aiming to characterize patterns of blank choice for the application of bipolar technology. For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et. al [-@fischer_macro_1984] was adopted (see Supplementary Materials for full list of variables). Studies by Sano [-@sano_functional_2012], and Gibaja et al. [-@gibaja_tecnologiy_2007] indicate that since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials). This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.

@@

Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use. Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had. For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch. Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution. On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight damaged platforms, and a higher frequency of scars on the hammered edge than on the opposed edge. The addition of other attributes drawing upon the work by Gonzalez-Urquijo & Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning. We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes. To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.

@@

A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% from Solutrean levels, as shown in (Table \@ref(tab:assemblage)). In terms of raw materials, the great majority of pieces were either made on quartz or chert, with chalcedony being represented only by `r round(techno_data %>% filter(RawMaterial == "Chalcedony") %>% summarise(n = n()), 2)` artifacts.

@@

As previously mentioned, concerning technological data, one of our main objectives was to characterize the choice of blanks involved in bipolar technology. Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site. Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material. For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)). In other periods, flakes were the preferred type of blank (Tables \@ref(tab:technosolutrean), \@ref(tab:technoproto) and \@ref(tab:technomagdalenian)). Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks. The small sample of Chalcedony blanks is exclusively represented by flakes.

@@

As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use. Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes (see Suplementary Materials for definition) cannot be straightforwardly performed. Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e. Length x Width) as an approximation for the overall dimensions of the artifacts. Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm. For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm. In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)). Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).

@@

d_platforms <- ggbarplot(n_axis, x = "RawMaterial", y = "perc",
          color = "DamagedPlatforms", fill = "DamagedPlatforms",
          palette = c("#00AFBB", "#E7B800", "#FC4E07", "#999999"),
          facet.by = "Chronology", xlab = "", ylab = "%")

@@

With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability. Similarly to what was registered for the technological attributes, most of the differences occur between raw materials rather than between chronologies. For this reason the results presented below focus only on overall and between raw materials variability. Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform. To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damage platforms are grouped into pairwise categories. When present, category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.

In three (Scar Arrangement, Extension and Edge Delineation) of the six represented variables the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable. A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`). The effect size statistic (Phi) suggests, however, a low practical significance for Scar Arrangement and a large pratical significance for Scar Faciality.

@@

To identify possible patterns of association among the qualitative variables used in our analysis \@ref(fig:morphoattributes), a Multiple Correspondence Analysis (MCA) was performed. In this analysis, raw material was used as a qualitative supplementary variable.

The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions. An inspection to the screeplot presented in Figure \@ref(fir:screeplot) suggests to restrict the analysis to the description of the first 4 dimensions. These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%). Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributes), and suggesting that patterns identified by previous studies [e.g. @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.

@@

```{r screeplot, echo = FALSE}

# Screeplot
fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "grey50", barcolor = "grey50")

@@

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, and elsewhere [@villa_neandertals_2018]. A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive. Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits. It is in the context of this latter point that bipolar technology may have played a major role. In Vale Boi, since the earliest occupations at c. 32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the aforementioned grease/marrow obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017]. For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture. In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision. The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.

In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for the onsite production and use of bone tools [@evora_osseous_2013]. According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood, and stone wedges are preferable than other types of material. For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics.

Importantly, our results show that the use of scale pieces at Vale Boi reveals techno-functional patterns that seem to stay fairly similar across all paleolithic horizons. One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes. One possible reason for these patterns might be the fact that quartz would be more easily available than chert, and thus tool economy would have been different between both materials. Another possibility for this is the difference in the physical properties of each raw material. While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it more often in order to continue obtaining usable edges. On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized. Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest, suggests that this hypothesis is likely incorrect. From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use. Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact. In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.

>>>>

---
title: "The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)"
author:
  - Pedro Horta (ICArEHB, University of Algarve, Faro, Portugal) pedroehorta@gmail.com
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
  - Nuno Bicho (ICArEHB, University of Algarve, Faro, Portugal) nbicho@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2: #subsistute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-socpsych-author-date.csl"
abstract: |
  Scaled or splintered pieces are one of the most common lithic artifacts type in Upper Paleolithic assemblages throughout Europe, especially in its westernmost regions. Despite this, even after one century of being identified there is still no consensus on how to define, analyze or even interpret these tools. In western Iberia there is a clear lack of comprehensive studies regarding this type of artifacts at a regional scale. In this paper we present a first techno-morphological analysis of a sample of scaled pieces from the Upper Paleolithic site of Vale Boi. Our first aim was to build upon existing analytical models in order to identify function and possible reduction strategies for these artifacts. Our second goal was to critically evaluate the role of these artifacts within western Iberia’s Upper Paleolithic.
  Our results showed that functional identification of scaled pieces is still not clear. By comparing our data with other author’s we found that current models could not be applied to the archaeological record, as the attribute variability is too high. Furthermore, in this region we found that higher frequencies of bipolar technology can be found related to residential sites due to both functional and cultural patterns. While we still cannot define a specific function for these artifacts (intermediate pieces or wedges for working hard raw materials or cores for the extraction of chips and small bladelets), it is clear that they had a major role in the variability and intensification of resource exploitation during the Upper Paleolithic in western Iberia.

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
keywords          : ""

@@

library(knitr)
library(tidyverse)
library(broom)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(ggplot2)
library(ggrepel)
library(maptools)
library(ggmap)
library(devtools)
library(legendMap)
library(tab)
library(sjstats)
library(powerAnalysis)
library(epitools)
library(wesanderson)
#library(ScaledPiecesVB)

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of portuguese paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces. Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g. @zilhao_o_1997], or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997], for example, argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.

@@

Vale Boi’s lithic studies have revealed a general tendency for stable technological and functional patterns throughout the Upper Paleolithic [e.g. @bicho_ledge:_2012; @bradtmoller_lithic_2016; @cascalheira_cross-scale_2017; @cascalheira_influencia_2013; @cascalheira_vale_2012; @gibaja_funcion_2011; @marreiros_as_2009; @marreiros_lithic_2013; @marreiros_lithic_2015; @marreiros_lithic_2018]. This can be explained, partially, by raw material availability. Most raw materials were procured locally, or regionally from deposits located at no more than 20 km away from the site [@bicho_paleolithique_2010]. 
Chert was the most used rock type for more complex retouched tools production, while quartz and greywacke were mostly used for flake extraction and simple retouched tools production. Other raw materials can be found at the site, but much more restricted, both diachronically and functionally, within each techno-complex. Schist, for instance, shows up in some occupations almost exclusively connected to mobile art. Dolerite is only  found in the Proto-Solutrean levels [@belmiro_o_2017], while chalcedony is mostly limited to the Proto-Solutrean and Solutrean levels [@cascalheira_vale_2012].

@@

We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends. These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed. As will become clear from our results, some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces. This detail has been previously referred by Villa et al. [-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.

Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes. In the first group, a series of variables traditionally used in lithic studies [e.g. @andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (see Supplementary Materials for full list of variables), aiming to characterize patterns of blank choice for the application of bipolar technology. For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et. al [-@fischer_macro_1984] was adopted (see Supplementary Materials for full list of variables). Studies by Sano [-@sano_functional_2012], and Gibaja et al. [-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials). This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.

@@

Macroscopic morpho-functional attributes were separately analyzed for each damaged platform, aiming to detect patterns of morphological change that occurred in artifacts during use. Following the work of de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011] and Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994], we expected that these attributes would be indicative of which function the artifacts had. For instance, according to de la Peña [-@de_la_pena_qualitative_2015; -@de_la_pena_sobre_2011], for pieces used as wedges the delineation of the damaged platforms are constantly asymmetrical and only the hammered edge clearly shows the typical *écaille* retouch. Furthermore, these pieces would have irregular shapes, variable scar size, and irregular scar distribution. On the other hand, pieces used as bipolar cores would have squared or rectangular shapes, symmetric straight damaged platforms, and a higher frequency of scars on the hammered edge than on the opposed edge. The addition of other attributes drawing upon the work by Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994] was made following the same reasoning. We expected, thus, to be able to differentiate pieces used as wedges from pieces used as cores, as both groups would show distinct combinations of attributes. To assist us with this differentiation we tested the presence of the referred patterns within our assemblage using descriptive and multivariate statistical analysis and comparing it with the data described by the referred authors.

@@

A total of `r summarise(techno_data, n())` scaled pieces were analyzed, of which `r round(techno_data %>% filter(Chronology == "Gravettian") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% come from Gravettian, and `r round(techno_data %>% filter(Chronology == "Solutrean") %>% summarise(n = n()) %>% mutate(n = (n / 139)*100), 2)`% from Solutrean levels, as shown in Table (\@ref(tab:assemblage)). In terms of raw materials, the great majority of pieces were either made on quartz or chert, with chalcedony being represented only by `r round(techno_data %>% filter(RawMaterial == "Chalcedony") %>% summarise(n = n()), 2)` artifacts.

@@

As previously mentioned, concerning technological data, one of our main objectives was to characterize the choice of blanks involved in bipolar technology. Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site. Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material. For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)). In other periods, flakes were the preferred type of blank (Tables \@ref(tab:technosolutrean), \@ref(tab:technoproto) and \@ref(tab:technomagdalenian)). Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks. The small sample of chalcedony blanks is exclusively represented by flakes.

@@

As in other classes of stone tools, scaled pieces' metric attributes are impacted by both the initial blank size as well as by the intensity of their use. Specifically, because some scaled pieces are used in multiple axes, direct comparisons for length and width of the typological axes cannot be straightforwardly performed. Since the majority of the analyzed pieces presented a rectangular outline, we use the area of a rectangle (i.e. Length x Width) as an approximation for the overall dimensions of the artifacts. Area calculations revealed a maximum of `r round(max(techno_data$Area, na.rm = TRUE), digits = 2)` mm and a minimum of `r round(min(techno_data$Area, na.rm = TRUE), digits = 2)` mm. For thickness the maximum is `r round(max(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm and the minimum is `r round(min(techno_data$Thickness, na.rm = TRUE), digits = 2)` mm. In general, mean values tend to be similar between raw materials, with some differences occurring within the Proto-Solutrean and Magdalenian assemblages, most certainly as a result of the small samples analysed for each of these periods (Figure \@ref(fig:meanplots)). Across techno-complexes, however, no significant statistical differences were detected when using an ANOVA test for both Area (*F* (`r aov_area$df[1]`, `r aov_area$df[2]`) = `r aov_area$statistic[1]`, *p* = `r aov_area$p.value[1]`, *d* = `r cohens_area$cohens.f`), and Thickness (*F* (`r aov_thickness$df[1]`, `r aov_thickness$df[2]`) = `r aov_thickness$statistic[1]`, *p* = `r aov_thickness$p.value[1]`, *d* = `r cohens_thickness$cohens.f`).

@@

d_platforms <- ggbarplot(n_axis, x = "RawMaterial", y = "perc",
          color = "DamagedPlatforms", fill = "DamagedPlatforms",
          facet.by = "Chronology", xlab = "", ylab = "%") + 
  scale_fill_manual(values = wes_palette("Rushmore")[c(1,3,4,5)]) +
  scale_color_manual(values=wes_palette("Rushmore")[c(1,3,4,5)])

@@

With few exceptions, most of the morpho-functional attributes show a fairly high degree of variability. Similarly to what was registered for the technological attributes, most of the differences occur between raw materials rather than between chronologies. For this reason the results presented below focus only on overall and between raw materials variability. Figure \@ref(fig:morphoattributes) shows the relative frequencies of all qualitative variables recorded for each damaged platform. To simplify representation, and to avoid wrong comparisons between hammered and active platforms, opposed damaged platforms were grouped into pairwise categories. When present, the category 'Other' represents the cluster of attributes which frequency was less than 10% within each variable.

In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable. A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`). The effect size statistic (Phi) suggests, however, a low practical significance for Scar Arrangement and a large pratical significance for Scar Faciality.

@@

To identify possible patterns of association among the qualitative variables used in our analysis, a Multiple Correspondence Analysis (MCA) was performed. In this analysis, raw material was used as a qualitative supplementary variable.

The first two dimensions of the MCA express 20.92% of the total dataset inertia, meaning that only that percentage of the total variability is explained by the plane combining the first two dimensions. An inspection to the screeplot presented in Figure \@ref(fig:screeplot) suggests restricting the analysis to the description of the first 4 dimensions. These dimensions present an amount of inertia slightly greater than that obtained by the 0.95-quantile of random distributions (37.74% against 34.37%). Still, it is a rather small percentage, somehow attesting the high variability suggested by our interpretation of Figure \@ref(fig:morphoattributes), and suggesting that patterns identified by previous studies [e.g. @de_la_pena_qualitative_2015; @de_la_pena_sobre_2011], in which certain combination of attributes were used to discriminate scaled pieces functionalities, are difficult to apply to the assemblage used in this study.

@@

```{r screeplot, echo = FALSE, fig.cap="Mutiple Correspondence Analysis screeplot."}

# Screeplot
fviz_screeplot(morpho_data_mca, addlabels = TRUE, ylim = c(0, 15), barfill = "#0B775E", barcolor = "#0B775E")

@@

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g. @villa_neandertals_2018]. A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive. Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits. It is in the context of this latter point that bipolar technology may have played a major role. In Vale Boi, since the earliest occupations at c. 32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the aforementioned grease/marrow obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017]. For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture. In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision. The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.

In addition, scaled pieces in Vale Boi may also be associated to bone tool production, given that Vale Boi is one of the portuguese Upper Paleolithic sites with significant evidence for onsite production and use of bone tools [@evora_osseous_2013]. According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood, and stone wedges are preferable than other types of material. For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their optimal morphological characteristics.

Importantly, our results show that the use of scale pieces at Vale Boi reveals techno-functional patterns that seem to stay fairly similar across all Upper Paleolithic horizons. One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes. One possible reason for these patterns might be the fact that quartz would be more easily available than chert, and thus tool economy would have been different between both materials. Another possibility for this is the difference in the physical properties of each raw material. While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it more often in order to continue obtaining usable edges. On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized. Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest, suggests that this hypothesis is likely incorrect. From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use. Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact. In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.

Several researchers argued that scaled pieces were used as cores for the extraction of chips or small bladelets [e.g. @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997]. Based on our data we find that this concept is somehow flawed, especially when applied to Vale Boi. First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al [-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm). Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, the functionality of chips of ~7mm is quite arguable, especially regarding the impact that these would have in a perforation or extension of a bleeding scenario, as opposed to using regular bladelets. Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with grooves that might be used for the insertion of stone implements. Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found. Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable. By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.

In Table \@ref(tab:scaledpiecesportugal) we present current data on Upper Paleolithic bipolar technology in Portugal. Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence. Moreover, it is clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first. It is also clear that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites. Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.  While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it should be a particularly high value.
